---
title: 'zzrctsim: Design, Novel Features, and Distributional Scope'
author: 'Ronald G. Thomas, Ph.D.'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
geometry: 'left=3cm,right=3cm,top=2cm,bottom=2cm'
header-includes:
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \setstretch{1.1}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 5, fig.height = 3.5, fig.align = 'center',
  dev = 'pdf', cache = TRUE, cache.path = 'cache/'
)
options(digits = 4, scipen = 1)
```

```{r load, include=FALSE, cache=FALSE}
# Deliberately not cached. pkgload::load_all() attaches the package,
# knitr records attached packages in cache/__packages, and on a later
# render it restores cached chunks with library(zzrctsim) -- which
# fails, because the package is loaded from source and never installed.
suppressMessages(pkgload::load_all(here::here(), quiet = TRUE))
```

# Purpose and scope

`zzrctsim` simulates randomized trials in order to evaluate design and
analysis choices. It exists because 21 of the 35 research compendia
under `~/prj/res` are simulation-dependent and 13 of them have no
simulation code written yet; a shared engine introduced now prevents
21 divergent reimplementations rather than reconciling them later.

This report has three parts. Section 2 sets out the architecture and
the contracts that hold it together. Sections 3 to 8 describe, in
detail, the six features that are not available in existing packages.
Sections 9 onward record the distributional scope: which response
families are supported, why the exponential family unifies less than
it appears to, which operations are Gaussian-only, and how the package
refuses work it cannot do.

Every measured figure quoted below was produced by running the code,
either in a chunk of this document or in the test suite, which
currently holds 194 assertions.

# Design

## The layered pipeline

The package is organized so that the code mirrors ADEMP, and a
manuscript's Methods subsections map onto modules rather than being
reconstructed from a monolithic script.

| Layer | Functions | Distribution-aware? |
|---|---|---|
| Design | `trial_schedule`, `runin_design` | no |
| Accrual | `accrue`, `close_out`, `realize_schedule` | no |
| DGM | `dgm_conditional`, `dgm_marginal`, `dgm_tte`, `cov_at` | yes |
| Generation | `generate_outcomes`, `generate_tte` | yes |
| Missingness | `dropout_mask`, `apply_mask`, `reference_based` | partly |
| Analysis | `fit_result`, `fit_ancova` | no |
| Driver | `run_simulation`, `sim_streams` | no |
| Performance | `compute_performance`, `nsim_for_mcse` | no |
| Power | `sim_power`, `power_curve`, `sample_size` | no |

Only three of nine layers know anything about the response
distribution. That asymmetry is the central design fact: the reuse
across endpoint families lives in the spine, and the variation lives
in the generator. It is why the endpoint families are best described
as *several generators behind one contract* rather than one generator
with a family switch.

## Three contracts

**The DGM object** answers `cov_at(dgm, times)` and nothing else is
assumed of it. Because it is a function of time rather than a fixed
matrix, a subject observed on an irregular grid, as staggered accrual
produces, is handled without special-casing.

**Complete data are generated first; missingness is a separate layer.**
This is what allows the information loss from dropout to be isolated by
comparison against the complete-data run, and it is the only way a
response-dependent mechanism can be expressed at all: a hazard that
reads `y` requires `y` to exist first.

**Fitters return one shape.** `fit_result()` carries `estimate`, `se`,
`p_value`, `ci_lower`, `ci_upper`, `df`, `converged`. The driver and
the performance measures never inspect a model object, so a new
analysis method is a function, not a change to the engine.

## What is deliberately not here

The package does not derive group-sequential boundaries (`rpact`,
`gsDesign` do), does not fit models beyond a reference ANCOVA
(`mmrm`, `nlme`, `lme4`, `survival` are the engines), does not perform
reference-based imputation for *analysis* (`rbmi` does), and is not a
general Monte Carlo harness (`SimDesign` is). It simulates trials and
measures what happens.

```{r shared-setup, include=FALSE}
# Objects used by examples throughout. Defined once here because the
# feature sections below precede the distributional sections that also
# use them.
s <- trial_schedule(treatment = 4, interval = 3)
arm <- factor(rep(c('placebo', 'active'), each = 300),
              levels = c('placebo', 'active'))
d <- runin_design(s, arm, reference = 'placebo')
ri <- function(t) matrix(1, length(t), 1)
b <- c(x_slope = 0.02, x_trt_active = -0.05)
```

# Novel feature 1: dual-parameterized data-generating mechanisms

A longitudinal DGM can be declared two ways: conditionally, through
random effects `(Z, G, sigma^2)`, or marginally, through a covariance
`V`. Existing packages pick one. `simr` and `lme4` are conditional
only; `mmrm` and `nlme` are marginal only. `zzrctsim` accepts either
and converts between them.

Forward conversion is exact and unremarkable: `V = Z G Z' + sigma^2 I`.
Backward conversion is the interesting direction, because it is not
always possible.

```{r dual}
tt <- 1:6
cs  <- cov_cs(sd = 1, rho = 0.5)(tt)
ar1 <- cov_ar1(sd = 1, rho = 0.6)(tt)
c(CS = certify(cs)$q_min, AR1 = certify(ar1)$q_min)
```

Compound symmetry is reachable from a single random effect; AR(1) at
six visits requires five. `as_conditional()` refuses an insufficient
budget rather than silently returning an approximation.

```{r dual_refuse, error=TRUE}
as_conditional(dgm_marginal(cov_ar1(1, 0.6)), tt, q = 2)
```

This matters because the relationship between the two
parameterizations is itself the subject of several compendia: 18 asks
when random-effects models reduce to simpler forms, 04 uses the
Woodbury identity to move between the two variance expressions, and
09 and 31 both turn on the same mapping.

# Novel feature 2: the reachability certificate

The refusal above is not a heuristic. A random-effects model with `q`
random effects and i.i.d. residuals induces `V = Z G Z' + sigma^2 I`,
where `Z G Z'` is positive semi-definite of rank at most `q`. A target
`V` is therefore reachable only if some `sigma^2 >= 0` makes
`V - sigma^2 I` positive semi-definite with rank at most `q`.

Subtracting `sigma^2 I` shifts every eigenvalue equally, so the
question is settled by the spectrum: take `sigma^2` to be the smallest
eigenvalue and count what remains.

```{r certificate}
ct <- certify(ar1)
c(q_min = ct$q_min, sigma2 = round(ct$sigma2, 4),
  max_reconstruction_error = max(abs(reconstruct(ct) - ar1)))
```

The certificate is **constructive**: the eigenvectors supply `Z`, the
shifted eigenvalues supply `G`, and the reconstruction is exact to
machine precision.

`reach_error()` converts "unreachable" into a magnitude, which is what
a practitioner actually needs:

```{r reacherr}
knitr::kable(reach_error(ar1), digits = 4, booktabs = TRUE,
             caption = 'Best attainable approximation to AR(1) at each
                        random-effects budget.')
```

A single random effect is wrong by 28% in relative Frobenius norm.
That is not a refinement of AR(1); it is a different covariance.

Two honest qualifications. First, `q_min` is **necessary but not
sufficient for a particular model**: the certificate lets `Z` be the
eigenvectors, whereas a real specification fixes `Z` to design columns
such as an intercept and a slope. It proves impossibility cleanly and
possibility only for an unconstrained `Z`. Second, the underlying
linear algebra is classical -- it is probabilistic PCA, equivalently
factor analysis with isotropic uniquenesses -- so the contribution is
packaging it as an automatic check in a simulation workflow, not the
mathematics.

# Novel feature 3: trial phase structure and staggered accrual

No surveyed package has a vocabulary for the phases of a trial.
`trial_schedule()` spans run-in, randomization, treatment and common
close, with run-in times negative and randomization at zero.

```{r schedule}
sch <- trial_schedule(run_in = 2, treatment = 4, common_close = 2,
                      interval = 3)
knitr::kable(as.data.frame(sch), booktabs = TRUE,
             caption = 'A three-phase schedule.')
```

Two design decisions are exposed rather than assumed.

**The common-close convention.** During the common close both arms are
off treatment. Whether the treated arm reverts to the control
trajectory or retains its accumulated benefit is a modeling choice
that changes the estimand, so both are offered:

```{r commonclose}
arm2 <- factor(c('placebo', 'active'), levels = c('placebo', 'active'))
rev <- runin_design(sch, arm2, common_close = 'revert')
ret <- runin_design(sch, arm2, common_close = 'retain')
rbind(revert = rev$x_trt_active[rev$arm == 'active'],
      retain = ret$x_trt_active[ret$arm == 'active'])
```

**The per-arm hinge.** Each arm independently either continues its
run-in slope through randomization or changes slope there. An unhinged
control arm constrains `beta = delta`, dropping one parameter, and is
what makes run-in observations directly informative about the control
slope -- the efficiency argument of compendium 04.

```{r hinge}
h  <- runin_design(sch, arm2, hinge = TRUE)
nh <- runin_design(sch, arm2,
                   hinge = c(placebo = FALSE, active = TRUE))
c(hinged_params = sum(grepl('^x_', names(h))),
  unhinged_params = sum(grepl('^x_', names(nh))))
```

The parameterization spans the same column space as the
`delta`/`beta`/`gamma` form used in compendium 04, so the closed-form
variance results derived there carry over unchanged.

**Staggered accrual with a common close-out.** Follow-up is not a
constant. Subjects enrol over an accrual window and the study ends on
one calendar date, so early enrollees are observed *longer* than the
nominal schedule and the last enrollee exactly as long.

```{r accrual}
set.seed(3)
s2 <- trial_schedule(run_in = 1, treatment = 4, interval = 3)
enr <- accrue(5, period = 12, pattern = 'linear')
cl  <- close_out(enr, s2, rule = 'lslv')
rz  <- realize_schedule(s2, enr, cl, extend = TRUE)
data.frame(enrolled = enr,
           visits = as.integer(table(rz$id)),
           follow_up = as.numeric(tapply(rz$time, rz$id, max)))
```

This is the mechanism that determines the information fraction at an
interim analysis, and it is why `cov_at()` must be a function of time.

# Novel feature 4: calibrated, mechanism-correct missingness

Dropout follows the selection-model factorization
`f(Y, R) = f(Y) f(R | Y)`, with the discrete-time hazard of Diggle and
Kenward (1994),

$$
  \mathrm{logit}\, P(\text{drop at } j)
   = \psi_0 + \psi_1 (y_{j-1} - c) + \psi_2 (y_j - c).
$$

Three properties distinguish this from the usual implementation.

**The mechanisms are nested restrictions of one model**, not three
separate models: MCAR is `psi1 = psi2 = 0`, MAR is `psi2 = 0`, MNAR
frees `psi2`. MNAR therefore *extends* MAR rather than replacing its
history dependence.

**The response is centred and `psi0` is calibrated** to a target
proportion, so the requested dropout is the realized dropout
regardless of `psi1` and `psi2`. Because the hazards depend only on
complete-data responses, the expected proportion has the closed form
`1 - prod(1 - p_j)` averaged over subjects, so calibration is a
one-dimensional root-find with no inner Monte Carlo.

```{r dropcal}
set.seed(9)
og <- generate_outcomes(d, dgm_conditional(G = diag(c(9, 0.04)),
                                           sigma2 = 4),
                        beta = c(x_slope = 0.5, x_trt_active = -0.25),
                        intercept = 20)
cal <- do.call(rbind, lapply(list(c(0, 0), c(0.15, 0), c(0.15, 0.10)),
  function(ps) do.call(rbind, lapply(c(0.10, 0.25, 0.50), function(tg) {
    z <- dropout_mask(og, target = tg, psi1 = ps[1], psi2 = ps[2])
    sp <- attr(z, 'spec')
    data.frame(mechanism = sp$mechanism, psi1 = ps[1], psi2 = ps[2],
               target = tg, realized = as.numeric(sp$expected))
  }))))
knitr::kable(cal, digits = 3, booktabs = TRUE, row.names = FALSE,
             caption = 'The target proportion is met across mechanisms
                        and across a range of psi1.')
```

Without centring, a hazard on an untransformed ADAS-Cog or CDR-SB
scale saturates: `plogis(qlogis(0.12) + 0.25 * 20) = 0.95`, and every
subject drops out at the first eligible visit while the nominal rate
still reads 0.12.

**Mask and filter are separate layers.** `dropout_mask()` generates a
pattern, `mask_from_data()` imports one from a completed trial, and
`apply_mask()` applies either. A mask is therefore a first-class
object that can be saved, inspected, and held fixed across replicates
under common random numbers -- none of which is possible when
generation and application are fused in one call.

# Novel feature 5: reference-based trajectories as a *generator*

`rbmi` implements reference-based assumptions for *analysis*.
`zzrctsim` implements them as *generators*, so that data whose true
post-discontinuation behaviour follows one assumption can be analysed
under another and the resulting bias measured.

```{r refgen}
set.seed(11)
mkr <- dropout_mask(og, target = 0.40)
shifts <- vapply(c('j2r', 'cir', 'lmcf', 'cr'), function(m) {
  z <- reference_based(og, mkr, method = m)
  post <- z$discontinued & og$arm == 'active'
  mean(z$mu[post] - og$mu[post])
}, numeric(1))
round(shifts, 3)
```

The ordering `0 < CIR < J2R` is the expected one: copying increments
in the reference retains the accumulated benefit that jumping to the
reference discards. Each subject's realized residual is preserved, so
the within-subject correlation survives the shift.

Discontinuation and missingness are deliberately separate operations,
because under a treatment-policy estimand post-discontinuation values
are observed and under a hypothetical estimand they are not. Fusing
them would make the treatment-policy case, which ICH E9(R1) puts
first, inexpressible.

# Novel feature 6: Monte Carlo discipline as infrastructure

Morris, White and Crowther (2019) is followed in the code, not only in
the write-up.

**Streams, not seeds.** `sim_streams()` produces `L'Ecuyer-CMRG`
substreams. Independence is guaranteed by construction, and the same
substreams replay across design cells, which is what common random
numbers require; per-replicate `set.seed(i)` gives neither guarantee
and desynchronizes silently when the amount of random-number
consumption changes between cells.

**Every replicate is reproducible in isolation.** The driver stores
the RNG state that produced each replicate, so a failing or anomalous
one can be re-run alone for diagnosis rather than by re-running the
study.

**Errors are recorded, not propagated.** A failure in generation or
analysis marks the replicate non-converged, retains the message, and
continues.

**Every performance measure carries a Monte Carlo standard error.**

```{r morris}
set.seed(5)
theta <- -0.25 * max(s$time)
res <- run_simulation(
  B = 200,
  generate = function() generate_outcomes(d,
    dgm_conditional(G = diag(c(9, 0.04)), sigma2 = 4),
    beta = c(x_slope = 0.5, x_trt_active = -0.25),
    intercept = 20),
  analyse = list(ancova = function(z) fit_ancova(z, reference = 'placebo')),
  estimand = estimand('final-visit contrast', theta), seed = 7L)
perf <- compute_performance(res)
knitr::kable(perf[, c('measure', 'estimate', 'mcse')], digits = 4,
             booktabs = TRUE, row.names = FALSE,
             caption = 'Morris Table 6 measures, each with its Monte
                        Carlo standard error.')
```

`nsim_for_mcse()` inverts these formulas so that the number of
replicates is justified rather than conventional.

# Sample size by simulation

`sample_size()` inverts a simulated power curve and then **confirms**
the selected size with an independent longer run, reporting the
achieved power and its Monte Carlo standard error. `simr`, the
established comparator, has no inverse: the analyst reads a
`powerCurve()` off by eye.

The practical value is that design features with no closed-form power
formula are priced directly. The same design, with and without 30%
MAR dropout:

```{r samplesize, cache=TRUE}
set.seed(2)
sch3 <- trial_schedule(treatment = 4, interval = 3)
gd3 <- dgm_conditional(G = diag(c(9, 0.04)), sigma2 = 4)
b3 <- c(x_slope = 0.5, x_trt_active = -0.12)
dfn <- function(n) runin_design(sch3,
  factor(rep(c('placebo', 'active'), each = n),
         levels = c('placebo', 'active')), reference = 'placebo')
common <- list(design_fn = dfn, dgm = gd3, beta = b3, intercept = 20,
  estimand = estimand('final contrast', b3[['x_trt_active']] * max(sch3$time)),
  analyse = list(ancova = function(z) fit_ancova(z, reference = 'placebo')),
  B = 300L, seed = 5L)
ss_c <- do.call(sample_size, c(list(target = 0.80,
  n_grid = c(40, 70, 110, 160), confirm_B = 800L), common))
ss_d <- do.call(sample_size, c(list(target = 0.80,
  n_grid = c(60, 110, 160, 220), confirm_B = 800L,
  dropout = list(target = 0.30, psi1 = 0.10)), common))
data.frame(scenario = c('complete data', '30% MAR dropout'),
           n_per_arm = c(ss_c$n_per_arm, ss_d$n_per_arm),
           confirmed_power = c(ss_c$confirmation$power,
                               ss_d$confirmation$power),
           mcse = c(ss_c$confirmation$mcse, ss_d$confirmation$mcse))
```

# What the exponential family does and does not unify

The exponential family unifies *univariate marginal* distributions.
It supplies no multivariate analogue: there is no canonical
multivariate binomial or multivariate Poisson corresponding to the
multivariate normal.

For a Gaussian response the marginal family and the dependence
structure are the same object, the covariance matrix. For every other
family they decouple, and a separate dependence mechanism must be
chosen. That choice, and not the marginal law, is the design decision.

`zzrctsim` takes the conditional route. Random effects are drawn,
`b_i ~ N(0, G)`, and the response is drawn conditionally,
`y_ij ~ family(g^{-1}(eta_ij))` with
`eta_ij = x_ij'beta + z_ij'b_i`. This is the GLMM construction used by
`lme4` and `glmmTMB`.

Two consequences follow, and both belong to the user rather than to
the package.

First, dependence between visits arises solely through the shared
`b_i`. The reachable dependence is exactly what the random-effects
structure can induce, and there is no marginal escape hatch:
`dgm_marginal()` has no meaning outside the Gaussian case.

Second, under a non-identity link the marginal mean is not
`g^{-1}(x'beta)`. The fixed effects are conditional on the random
effects, so conditional and marginal treatment effects differ in
magnitude. An `estimand()` used with a non-Gaussian family must state
which of the two it names.

# Supported families

```{r families}
s <- trial_schedule(treatment = 4, interval = 3)
arm <- factor(rep(c('placebo', 'active'), each = 300),
              levels = c('placebo', 'active'))
d <- runin_design(s, arm, reference = 'placebo')
ri <- function(t) matrix(1, length(t), 1)
b <- c(x_slope = 0.02, x_trt_active = -0.05)

specs <- list(
  list(nm = 'gaussian(identity)', f = gaussian(),         s2 = 4,    ic = 20),
  list(nm = 'binomial(logit)',    f = binomial(),         s2 = NULL, ic = 0),
  list(nm = 'binomial(probit)',   f = binomial('probit'), s2 = NULL, ic = 0),
  list(nm = 'poisson(log)',       f = poisson(),          s2 = NULL, ic = 1)
)

set.seed(8)
out <- do.call(rbind, lapply(specs, function(sp) {
  g <- dgm_conditional(G = matrix(0.5), sigma2 = sp$s2,
                       z = ri, family = sp$f)
  o <- generate_outcomes(d, g, beta = b, intercept = sp$ic)
  w <- matrix(o$y, ncol = nrow(s), byrow = TRUE)
  data.frame(family = sp$nm, min = min(o$y), max = max(o$y),
             mean = mean(o$y), within_cor = cor(w[, 2], w[, 3]))
}))
knitr::kable(out, digits = 3, booktabs = TRUE,
             caption = 'Response support and within-subject
                        dependence by family.')
```

The within-subject correlation is induced by the shared random
intercept in every case. It is not a free parameter: it is whatever
the random-effects structure implies on the response scale, and it
differs by family even with `G` held fixed.

Negative binomial responses require `theta`, the dispersion in the
`size` parameterization of `rnbinom()`. Binomial responses accept
`trials`, defaulting to `1` for a binary outcome.

# Time-to-event: single and recurrent

Time-to-event does not fit the covariance-based parameterization of
the previous section, and this is a structural fact rather than an
omission. Censoring, hazards and risk sets are a different kind of
object from a marginal law, so `dgm_tte()` is a sibling generator
behind the same contract rather than another `family` argument.

Generation is by **inversion of the cumulative hazard**, which is
exact rather than a discretization of a grid, and works identically
for both baseline hazards described below. Under proportional
hazards the individual cumulative hazard is
`H_i(t) = H_0(t) exp(lp_i)`, so if `E ~ Exp(1)` then
`T = H_0^{-1}(E / exp(lp_i))` has exactly the intended distribution.
For recurrent events the same inversion is applied to successive
partial sums of i.i.d. `Exp(1)` variates, which yields the event times
of a non-homogeneous Poisson process with intensity `h_i(t)`.

## Single event

```{r tte_single}
set.seed(4001)
subj <- data.frame(id = 1:4000, trt = rep(0:1, each = 2000))
g_wb <- dgm_tte(shape = 1.6, scale = 15)
g_wb
ev <- generate_tte(subj, g_wb, beta = c(trt = log(0.65)),
                   followup = 30)
head(ev, 3)
```

Administrative censoring is the natural companion to staggered
accrual: passing `followup = close - enroll` censors every subject at
the common close-out date, with follow-up decreasing in enrolment
time.

```{r tte_recover}
fit_cox <- survival::coxph(survival::Surv(time, status) ~ trt,
                           data = ev)
c(true_log_HR = log(0.65),
  estimated = unname(coef(fit_cox)),
  se = sqrt(unname(vcov(fit_cox)[1, 1])),
  censored_fraction = mean(ev$status == 0))
```

## Recurrent events

Recurrent events are returned in counting-process format -- one row
per interval with `tstart`, `tstop`, `status` and `enum` -- which is
what `survival::coxph()` expects for an Andersen-Gill model. The
intervals for a subject tile the follow-up window contiguously and end
with a censored row.

```{r tte_recurrent}
set.seed(4002)
subj2 <- data.frame(id = 1:1500, trt = rep(0:1, each = 750))
g_rec <- dgm_tte(shape = 1, scale = 10, recurrent = TRUE)
rec <- generate_tte(subj2, g_rec, beta = c(trt = 0), followup = 20)
head(rec, 4)
```

A Gaussian **frailty** on the log-hazard scale induces within-subject
dependence between gap times, which is what separates a recurrent
process from independent replications. Its signature is
overdispersion:

```{r tte_frailty}
set.seed(4003)
counts <- function(sd) {
  g <- dgm_tte(shape = 1, scale = 10, recurrent = TRUE,
               frailty_sd = sd)
  r <- generate_tte(subj2, g, beta = c(trt = 0), followup = 20)
  unlist(tapply(r$status, r$id, sum))
}
data.frame(frailty_sd = c(0, 0.8),
           mean_events = c(mean(counts(0)), mean(counts(0.8))),
           var_events  = c(var(counts(0)),  var(counts(0.8))))
```

Without frailty the counts are Poisson, variance equal to mean. With
frailty they are overdispersed. Note the consequence for
interpretation: a frailty makes the *marginal* distribution
non-proportional even though the conditional model is proportional
hazards, which is the usual reason a fitted hazard ratio attenuates
toward one. The estimand must say whether it names the conditional or
the marginal hazard ratio.

## Baseline hazards, and departures from proportionality

Two baseline hazards are available. The Weibull covers monotone
hazards and reduces to the exponential at `shape = 1`. The
**piecewise exponential** expresses an arbitrary hazard shape without
committing to a parametric family, which is why it is the usual
choice in trial simulation and what `simtrial` is built on. Its
cumulative hazard is piecewise linear, so inversion remains exact.

```{r tte_pw}
g_pw <- dgm_tte(breaks = c(6, 12), rates = c(0.02, 0.06, 0.15))
g_pw
set.seed(4101)
sp <- data.frame(id = 1:40000, trt = 0)
e_pw <- generate_tte(sp, g_pw, beta = c(trt = 0), followup = 20)
emp <- function(lo, hi) {
  sum(e_pw$status == 1L & e_pw$time > lo & e_pw$time <= hi) /
    sum(pmax(0, pmin(e_pw$time, hi) - lo))
}
data.frame(interval = c('[0,6)', '[6,12)', '[12,20]'),
           specified = c(0.02, 0.06, 0.15),
           realized = c(emp(0, 6), emp(6, 12), emp(12, 20)))
```

**A delayed treatment effect** is the more consequential addition,
because it breaks proportional hazards on purpose. Setting
`effect_lag = L` gives intensity
`h_i(t) = h_0(t) exp(lp_i * I(t > L))`: the arms are indistinguishable
until `L` and diverge thereafter. The individual cumulative hazard
remains invertible in closed form, so this is still generated exactly
rather than by discretization.

This shape is standard in immuno-oncology and is what anti-amyloid
trials in Alzheimer's disease are generally argued to show. A constant
hazard ratio cannot represent it.

```{r tte_lag}
set.seed(4102)
s4 <- data.frame(id = 1:12000, trt = rep(0:1, each = 6000))
lag <- 8
e_lag <- generate_tte(s4, dgm_tte(shape = 1, scale = 20,
                                  effect_lag = lag),
                      beta = c(trt = log(0.4)), followup = 40)
fz <- survival::coxph(survival::Surv(time, status) ~ trt, data = e_lag)
data.frame(
  true_post_lag_HR = 0.4,
  fitted_constant_HR = exp(unname(coef(fz))),
  ph_test_p = survival::cox.zph(fz)$table[1, 'p'])
```

The fitted constant hazard ratio is attenuated toward one relative to
the true post-lag ratio of 0.4, and `cox.zph()` rejects
proportionality. Both are the intended behaviour: the generator
produces data that a proportional-hazards summary misrepresents, which
is exactly what is needed to study how badly it misrepresents them.

## An endpoint derived from the longitudinal trajectory

Compendium 13 compares MMRM against survival analysis. That comparison
is only meaningful if both endpoints describe the *same* underlying
process, rather than being drawn independently.
`tte_from_trajectory()` defines the event as the first visit at which
a generated trajectory crosses a threshold.

```{r tte_derived}
set.seed(4005)
s6 <- trial_schedule(treatment = 6, interval = 3)
arm6 <- factor(rep(c('placebo', 'active'), each = 300),
               levels = c('placebo', 'active'))
d6 <- runin_design(s6, arm6, reference = 'placebo')
lo <- generate_outcomes(d6, dgm_conditional(G = diag(c(4, 0.02)),
                                            sigma2 = 2),
                        beta = c(x_slope = 0.35, x_trt_active = -0.20),
                        intercept = 10)
evd <- tte_from_trajectory(lo, threshold = 16, direction = 'above')
a6 <- arm6[evd$id]
fit_d <- survival::coxph(survival::Surv(time, status) ~ a6, data = evd)
data.frame(arm = c('placebo', 'active'),
           crossed = c(mean(evd$status[a6 == 'placebo']),
                       mean(evd$status[a6 == 'active'])),
           mean_time_given_crossed = c(
             mean(evd$time[evd$status == 1 & a6 == 'placebo']),
             mean(evd$time[evd$status == 1 & a6 == 'active'])))
```

This table is worth pausing on. The treatment effect is large -- 64%
of placebo cross against 25% of active, a Cox hazard ratio of
`r round(exp(unname(coef(fit_d))), 3)` -- yet the mean crossing time
*among those who crossed* is nearly identical between arms.
Conditioning on having crossed selects the fastest progressors within
each arm, and selects more severely in the arm where crossing is
rarer. It is a compact illustration of why a survival endpoint is
analysed through the hazard with censored observations retained,
rather than through the mean event time of the subset that had events.

Subjects who never cross are censored at their last observed visit, so
dropout and administrative censoring both propagate correctly into the
survival endpoint.

# Three scales, all retained

The GLMM path returns three columns rather than collapsing them,
because for a non-identity link they are genuinely different
quantities and a reader cannot reconstruct one from another without
knowing the link.

```{r scales}
set.seed(21)
gb <- dgm_conditional(G = matrix(0.5), z = ri, family = binomial())
ob <- generate_outcomes(d, gb, beta = b, intercept = 0)
knitr::kable(head(ob[, c('id', 'time', 'mu', 'eta', 'cmean', 'y')], 6),
             digits = 3, booktabs = TRUE,
             caption = 'mu is the fixed-effect linear predictor, eta
                        adds the random effects, cmean applies the
                        inverse link, y is the draw.')
```

The realized random effects are retained on a `ranef` attribute, so a
replicate can be interrogated rather than only re-run.

# What is Gaussian-only, and why

Four operations presuppose a Gaussian response. Each is now guarded.

## Covariance of the response

`cov_at()` returns the covariance of the response. For a binomial or
Poisson response the variance is determined by the mean, so it cannot
be specified separately, and the function refuses.

```{r covat, error=TRUE}
cov_at(gb, s$time)
```

`linpred_cov()` is the defined alternative for every family: it
returns `Z G Z'`, the covariance induced on the linear-predictor
scale.

```{r linpred}
linpred_cov(gb, s$time)[1:3, 1:3]
```

## Conversion and the reachability certificate

`certify()`, `as_conditional()` and `as_marginal()` are results about
`V = Z G Z' + sigma^2 I`. They are Gaussian statements and have no
non-Gaussian analogue, because the marginal parameterization they
convert to does not exist outside that case.

## Reference-based post-discontinuation trajectories

`reference_based()` preserves each subject's additive residual
`y - mu` while shifting the mean. On a count or binary scale that
construction produces values outside the support: a probability
shifted by a residual is not a probability, and a shifted count need
not be an integer. The function refuses rather than returning
impossible data.

```{r refbased, error=TRUE}
mk <- dropout_mask(ob, target = 0.30)
reference_based(ob, mk, method = 'j2r')
```

## Response-dependent missingness

The dropout hazard follows Diggle and Kenward (1994),

$$
  \mathrm{logit}\, P(\text{drop at } j)
   = \psi_0 + \psi_1 (y_{j-1} - c) + \psi_2 (y_j - c),
$$

with the response centred at `c`. Two things are then scale-dependent:
the default centring, the mean of `y`; and the interpretation of
`psi1` as a change in log-odds per unit of `y`.

The guard is deliberately narrow. **MCAR is permitted for every
family**, because its hazard never reads `y` at all. Only
response-dependent missingness is restricted, and it is permitted for
a non-Gaussian family as soon as `center` is supplied explicitly,
since supplying it is the act of deliberation the guard exists to
force.

```{r dropguard}
guard <- function(label, expr) {
  r <- tryCatch({ expr; 'allowed' },
                error = function(e) 'blocked')
  data.frame(call = label, result = r)
}
knitr::kable(rbind(
  guard('binomial, MCAR',              dropout_mask(ob, target = 0.2)),
  guard('binomial, MAR, no center',    dropout_mask(ob, target = 0.2,
                                                    psi1 = 0.3)),
  guard('binomial, MAR, center given', dropout_mask(ob, target = 0.2,
                                                    psi1 = 0.3,
                                                    center = 0.5)),
  guard('binomial, reference_based',   reference_based(ob, mk, 'j2r'))
), booktabs = TRUE, row.names = FALSE,
   caption = 'Guards permit MCAR for any family and require an
              explicit centring for response-dependent mechanisms.')
```

# What is family-agnostic

The guards are narrow because most of the package makes no
distributional assumption at all. The design, missingness-application,
driver and performance layers operate on estimates and indices, not on
a response distribution:

| Layer | Functions | Family-agnostic |
|---|---|---|
| Schedule | `trial_schedule`, `runin_design` | yes |
| Accrual | `accrue`, `close_out`, `realize_schedule` | yes |
| Mask application | `apply_mask`, `mask_from_data` | yes |
| Driver | `run_simulation`, `sim_streams`, `with_rng_state` | yes |
| Performance | `compute_performance`, `nsim_for_mcse` | yes |
| Power | `sim_power`, `power_curve`, `sample_size` | yes |
| Generation | `generate_outcomes`, `generate_tte` | family-aware |
| Covariance | `cov_at`, `certify`, `as_conditional` | Gaussian only |
| Mask generation | `dropout_mask`, `apply_dropout` | Gaussian, or explicit centring |
| Pattern mixture | `reference_based` | Gaussian only |

This is why the endpoint families of the design document are described
as *five generators behind one contract* rather than one generator
with five settings. The reuse lives in the spine, which is built; the
variation lives in the generator, which is not yet complete.

The spine carries a non-Gaussian study end to end today:

```{r endtoend}
fit_glm <- function(z) {
  zz <- z[abs(z$time - max(z$time)) < 1e-8, ]
  m <- tryCatch(glm(y ~ arm, data = zz, family = binomial()),
                error = function(e) NULL)
  if (is.null(m)) return(null_fit())
  cf <- summary(m)$coefficients
  fit_result(cf[2, 'Estimate'], cf[2, 'Std. Error'], cf[2, 'Pr(>|z|)'])
}
res <- run_simulation(
  B = 300,
  generate = function() generate_outcomes(d, gb, beta = b, intercept = 0),
  analyse = list(glm = fit_glm),
  estimand = estimand('conditional log-OR at final visit', NA_real_),
  seed = 2L
)
perf <- compute_performance(res, theta = 0)
knitr::kable(perf[perf$measure %in% c('n_converged', 'emp_se',
                                      'mod_se', 'rejection'),
                  c('measure', 'estimate', 'mcse')],
             digits = 4, booktabs = TRUE, row.names = FALSE,
             caption = 'Morris Table 6 measures for a binary endpoint.')
```

Note that `theta` is passed as `0` for a null check rather than as a
marginal log-odds ratio. The estimand under a logit link is ambiguous
between the conditional and marginal effect, and the package does not
resolve that ambiguity on the user's behalf.

# Boundaries and open work

Time-to-event, added above, is one of the two families that fall
outside the exponential family and needs its own generator rather than
another `family` argument. One remains.

**Mixtures of alpha-stable laws.** No closed-form density, and for
stability index below two no finite variance. That last point reaches
beyond the generator: without moments, bias, empirical standard error
and mean squared error are undefined, so `compute_performance()` must
dispatch on whether moments exist rather than return a finite-looking
number. This is the one place where an endpoint family breaks the
family-agnostic spine.

A copula route would allow arbitrary marginals with a specified
correlation, but the requested correlation is not the achieved one for
discrete responses, and the shortfall worsens as events become rarer.
If that route is added, the achieved correlation must be computed and
returned rather than assumed.

# Reproducibility

```{r session, echo=FALSE}
cat('zzrctsim: development version\n')
cat('R: ', R.version.string, '\n', sep = '')
```

```{r footer, echo=FALSE, results='asis'}
local({
  d <- dirname(knitr::current_input(dir = TRUE))
  while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) &&
         d != dirname(d)) d <- dirname(d)
  source(file.path(d, 'tools', 'stamp-render.R'), local = TRUE)
  footer_stamp()
})
```
