---
title: "zzrctsim and simr: two routes to simulated power"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zzrctsim and simr: two routes to simulated power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzrctsim)
has_simr <- requireNamespace("simr", quietly = TRUE)
```

```{r notice, echo = FALSE, results = "asis", eval = !has_simr}
cat("**Note.** `simr` is not installed in the library used to build",
    "this vignette, so its chunks are shown but not run. Install it",
    "with `install.packages('simr')` to execute them. The `zzrctsim`",
    "chunks run either way.\n")
```

## Scope

`simr` (Green and MacLeod, 2016) is the established package for
simulation-based power analysis of mixed models, and it is the natural
comparator. The two tools solve adjacent problems, and the honest
summary is that neither dominates: `simr` is better when you have a
pilot model, `zzrctsim` is better when the trial design or the
covariance structure is the object of study.

## The same question, both ways

Take a two-arm trial, four post-randomization visits, a random
intercept, and a treatment effect on the rate of change. How many
subjects per arm for 80% power?

### simr

```{r simr-setup, eval = has_simr}
library(simr)
n0 <- 60
X <- expand.grid(visit = 1:4, subject = factor(1:(2 * n0)))
X$arm <- factor(rep(c("placebo", "active"), each = 4 * n0))

m <- makeLmer(y ~ visit * arm + (1 | subject),
              fixef = c(20, 0.5, 0, -0.25),
              VarCorr = 9, sigma = 2,
              data = X)
pc <- powerCurve(m, test = fixed("visit:arm", "t"),
                 along = "subject",
                 breaks = c(40, 80, 120, 160), nsim = 200)
print(pc)
```

`powerCurve()` returns power at each break with a binomial confidence
interval. There is no inverse: you read the crossing off by eye, or
interpolate by hand.

### zzrctsim

```{r zz-setup}
s <- trial_schedule(treatment = 4, interval = 1)
g <- dgm_conditional(G = matrix(9), sigma2 = 4,
                     z = function(t) matrix(1, length(t), 1))
b <- c(x_slope = 0.5, x_trt_active = -0.25)
design_fn <- function(n) {
  runin_design(s, factor(rep(c("placebo", "active"), each = n),
                         levels = c("placebo", "active")),
               reference = "placebo")
}
ss <- sample_size(
  target = 0.80, n_grid = c(20, 40, 80, 120), confirm_B = 600L,
  design_fn = design_fn, dgm = g, beta = b, intercept = 20,
  estimand = estimand("final contrast",
                      b[["x_trt_active"]] * max(s$time)),
  analyze = list(ancova = function(z)
    fit_ancova(z, reference = "placebo")),
  B = 300L, seed = 5L)
ss
```

The curve is inverted and the selected size **confirmed** by an
independent longer run, which reports the achieved power with its
Monte Carlo standard error. That confirmation is the number to quote:
a size read off an interpolated curve inherits the curve's Monte Carlo
error, and the confirmation run measures it directly.

## Where the two diverge

### Covariance

`simr` takes `VarCorr`, the random-effects covariance. The reachable
marginal covariance is therefore `Z G Z' + sigma^2 I`. A random
intercept gives compound symmetry; nothing else is available.

```{r reachable}
V_ri <- cov_at(g, s$time)
round(V_ri[1, ], 3)

sds <- c(4.5, 4.5, 5.4, 6.3, 7.2)
certify(cov_ar1(sd = sds, rho = 0.6)(1:5))$q_min
```

An empirically estimated covariance with heterogeneous variances needs
four random effects for five visits. It is not expressible through
`VarCorr` at any parameter setting, so a study whose subject *is* the
covariance cannot be run in `simr`.

### Dropout

This is the sharper difference. `simr` generates the response from a
fixed design matrix, so missingness has to be imposed on the design
*before* generation. It can therefore be MCAR or covariate-dependent,
but it cannot depend on the response: MAR conditioning on an earlier
observed value, and MNAR, are structurally unreachable.

`zzrctsim` generates complete data first and applies missingness
afterwards, which is what makes response-dependent mechanisms
expressible:

```{r dropout}
d <- design_fn(100)
dat <- generate_outcomes(d, g, beta = b, intercept = 20)
mk <- dropout_mask(dat, target = 0.30, psi1 = 0.15)
attr(mk, "spec")$mechanism
```

For a study about dropout adjustment, that is decisive.

### Design vocabulary

`simr` has `extend(along =, within =, n =)`, a clean abstraction for
growing a design along any factor. It has no notion of a run-in phase,
a common close, staggered accrual, or an interim analysis.
`zzrctsim` has those and no `extend()`.

```{r phases}
sch <- trial_schedule(run_in = 2, treatment = 4, common_close = 2,
                      interval = 3)
table(sch$phase)
```

## Where simr is the better choice

Stated plainly, because a comparison the local package wins on every
axis is not a comparison.

**You have pilot data.** `simr` takes a fitted `merMod` directly. No
covariance has to be elicited, no parameters guessed — the model you
already fitted *is* the specification. `zzrctsim` has no equivalent
path and would require you to read the variance components off and
retype them.

**Your outcome is a GLMM.** `simr` handles binomial and Poisson
mixed models today, including tests via likelihood ratio and
parametric bootstrap. `zzrctsim` generates those families but ships
only one reference fitter.

**You want an established tool.** `simr` is widely used and
recognized by reviewers. A local package carries the burden of proof,
and for a routine power calculation that burden is not worth carrying.

**Your design is not a clinical trial.** `simr` came from ecology and
is general across factorial and crossed designs. `zzrctsim`'s
vocabulary — arms, visits, dropout, interim looks — is narrow by
intent.

## Using both

They compose. A pilot model in `simr` gives a starting sample size
quickly; `zzrctsim` then prices the design features `simr` cannot
express, starting from that number rather than from a guess.

```{r both, eval = FALSE}
# 1. simr, from a fitted pilot model: a starting n.
# 2. zzrctsim, from that n: what does 30% MAR dropout cost?
ss_drop <- sample_size(
  target = 0.80, n_grid = c(60, 110, 160, 220), confirm_B = 600L,
  design_fn = design_fn, dgm = g, beta = b, intercept = 20,
  estimand = estimand("final contrast",
                      b[["x_trt_active"]] * max(s$time)),
  analyze = list(ancova = function(z)
    fit_ancova(z, reference = "placebo")),
  dropout = list(target = 0.30, psi1 = 0.10),
  B = 300L, seed = 5L)
ss_drop$n_per_arm
```

## Summary

| | `simr` | `zzrctsim` |
|---|---|---|
| Starting point | fitted `merMod`, or `makeLmer` | design plus DGM specification |
| Covariance | `VarCorr`, so `Z G Z' + sigma^2 I` | either parameterization, any `V` |
| Sample size | `powerCurve()` read off by hand | inverted and confirmed |
| Dropout | design-determined only | MCAR, MAR, MNAR, calibrated |
| Run-in, common close, accrual | none | native |
| Non-Gaussian | GLMMs with several tests | generation only, one fitter |
| Output | rejection rate with an interval | Morris Table 6, MCSE throughout |
| RNG | `set.seed` | `L'Ecuyer-CMRG` substreams, per-replicate state |

Use `simr` where a fitted pilot model is the natural starting point
and the outcome is generalized. Use `zzrctsim` where the covariance
structure, the dropout mechanism, or the trial phase structure is what
you are studying.
