---
title: "zzrctsim and lme4: generating, and fitting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zzrctsim and lme4: generating, and fitting}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzrctsim)
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
```

```{r skip, echo = FALSE, results = "asis", eval = !has_lme4}
cat("**`lme4` is not installed, so the code below is not evaluated.**\n")
```

## Two different relationships

`lme4` appears twice in a simulation study, and the two roles should
not be confused.

As a **generator**, it competes with `zzrctsim`: `simulate()` has a
formula method that produces data from a specification with no fitted
model. As a **fitter**, it complements `zzrctsim`: `lmer()` is one of
the analysis methods you would want to evaluate.

The first comparison has a clear answer. The second is not a
comparison at all — it is how the two are meant to be used together.

## lme4 as a generator

`lme4::simulate()` can generate without any data, given a formula and
a parameter list.

```{r lme4-gen, eval = has_lme4}
library(lme4)
set.seed(1)
n_per_arm <- 100; p <- 5
X <- expand.grid(visit = 1:p, subject = factor(1:(2 * n_per_arm)))
X$arm <- factor(rep(c("placebo", "active"), each = p * n_per_arm))

sd_int <- 3; sigma <- 2.4
y <- simulate(~ visit * arm + (1 | subject),
              newdata = X,
              newparams = list(beta = c(20, 0.5, 0, -0.25),
                               theta = sd_int / sigma,
                               sigma = sigma),
              family = gaussian, nsim = 1)[[1]]
length(y)
```

That works, and for a random-intercept model it is straightforward.
Two things about it are worth knowing.

### theta is not the covariance

`theta` is the lower triangle, in column-major order, of the Cholesky
factor of the random-effects covariance **divided by the residual
standard deviation**. For one random intercept it collapses to
`sd / sigma`, which is easy. For a correlated intercept and slope it
does not:

```{r theta, eval = has_lme4}
G <- matrix(c(9, 0.6 * 3 * 0.2, 0.6 * 3 * 0.2, 0.04), 2, 2)
sigma <- 2
theta <- (t(chol(G)) / sigma)[lower.tri(diag(2), diag = TRUE)]
round(theta, 4)

# Round-trip, to confirm the ordering:
Lam <- matrix(0, 2, 2); Lam[lower.tri(Lam, diag = TRUE)] <- theta
max(abs(sigma^2 * Lam %*% t(Lam) - G))
```

Getting the ordering wrong produces a *different covariance*, silently.
`zzrctsim` takes `G` on the natural variance scale:

```{r zz-gen, eval = has_lme4}
g <- dgm_conditional(G = G, sigma2 = sigma^2)
round(cov_at(g, 1:3), 3)
```

### The reachable covariance is constrained

This is the more consequential difference. Because `lme4` generates
from random effects, the marginal covariance is whatever
`Z G Z' + sigma^2 I` implies. A random intercept gives compound
symmetry and nothing else:

```{r reachable, eval = has_lme4}
ri <- dgm_conditional(G = matrix(9), sigma2 = 4,
                      z = function(t) matrix(1, length(t), 1))
V <- cov_at(ri, 1:5)
round(V[1, ], 3)
```

Constant off-diagonal. Suppose instead you estimated this from a
completed trial — variances rising over follow-up, correlations
decaying with lag:

```{r unreachable, eval = has_lme4}
sds <- c(4.5, 4.5, 5.4, 6.3, 7.2)
V_emp <- cov_ar1(sd = sds, rho = 0.6)(1:5)
certify(V_emp)$q_min
```

Four random effects for five visits. A random intercept-and-slope
model cannot represent this covariance at any parameter setting, so it
cannot generate from it either. `zzrctsim` can, because the marginal
route needs no random-effects representation:

```{r marginal-gen, eval = has_lme4}
s <- trial_schedule(treatment = 4, interval = 1)
arm <- factor(rep(c("placebo", "active"), each = 150),
              levels = c("placebo", "active"))
d <- runin_design(s, arm, reference = "placebo")
gm <- dgm_marginal(cov_ar1(sd = sds, rho = 0.6))
dat <- generate_outcomes(d, gm,
                         beta = c(x_slope = 0.5, x_trt_active = -0.25),
                         intercept = 20)
round(cor(matrix(dat$y, ncol = nrow(s), byrow = TRUE))[1, ], 3)
```

The correlations decay, which no random-intercept generator produces.

## The two agree where they overlap

Where `lme4` *can* express a mechanism, the two generators agree. Both
are drawing from the same multivariate normal.

```{r agree, eval = has_lme4}
set.seed(99)
G1 <- matrix(9); sig <- 2
zz <- dgm_conditional(G = G1, sigma2 = sig^2,
                      z = function(t) matrix(1, length(t), 1))
target <- cov_at(zz, 1:5)

# lme4, same specification
Xa <- expand.grid(visit = 1:5, subject = factor(1:4000))
ya <- simulate(~ 1 + (1 | subject), newdata = Xa,
               newparams = list(beta = 0, theta = sqrt(9) / sig,
                                sigma = sig),
               family = gaussian, nsim = 1)[[1]]
emp_lme4 <- cov(matrix(ya, ncol = 5, byrow = TRUE))

round(rbind(target = target[1, ], lme4 = emp_lme4[1, ]), 3)
```

## lme4 as a fitter

This is the productive relationship. `zzrctsim` does not fit models —
it defines a contract and drives the loop. Wrapping `lmer()` takes
five lines.

```{r fitter, eval = has_lme4}
fit_lmer_slope <- function(z) {
  m <- try(lmer(y ~ time * arm + (1 | id), data = z,
                REML = TRUE), silent = TRUE)
  if (inherits(m, "try-error")) return(null_fit())
  cf <- summary(m)$coefficients
  trm <- grep("^time:arm", rownames(cf), value = TRUE)[1]
  fit_result(estimate = cf[trm, "Estimate"],
             se = cf[trm, "Std. Error"],
             p_value = 2 * pnorm(-abs(cf[trm, "t value"])))
}
```

Now compare it against ANCOVA on the *same* generated data, which is
the paired comparison Morris et al. recommend:

```{r compare-fitters, eval = has_lme4}
set.seed(7)
s2 <- trial_schedule(treatment = 4, interval = 3)
arm2 <- factor(rep(c("placebo", "active"), each = 80),
               levels = c("placebo", "active"))
d2 <- runin_design(s2, arm2, reference = "placebo")
g2 <- dgm_conditional(G = diag(c(9, 0.04)), sigma2 = 4)
b2 <- c(x_slope = 0.5, x_trt_active = -0.25)

res <- run_simulation(
  B = 100,
  generate = function() generate_outcomes(d2, g2, beta = b2,
                                          intercept = 20),
  analyze = list(
    ancova = function(z) fit_ancova(z, reference = "placebo"),
    lmer_slope = fit_lmer_slope),
  estimand = estimand("see note", NA_real_),
  seed = 21L)
perf <- compute_performance(res, theta = 0)
perf[perf$measure %in% c("n_converged", "emp_se", "mod_se",
                         "rel_error_mod_se"), ]
```

The estimand is deliberately left as `NA` and `theta = 0` passed for
the summary, because **the two fitters target different quantities**:
ANCOVA estimates the contrast at the final visit, `lmer` with a
`time:arm` term estimates a difference in slopes. Comparing their bias
against a common `theta` would be meaningless. Their empirical
standard errors and convergence rates are comparable; their point
estimates are not.

That is a general caution rather than a limitation of either package:
a simulation comparing methods must first check they estimate the same
thing.

## Summary

| | `lme4` | `zzrctsim` |
|---|---|---|
| Generate without data | yes, `simulate()` formula method | yes |
| Covariance specification | `theta`, a scaled Cholesky | `G` on the variance scale, or `V` directly |
| Reachable covariance | `Z G Z' + sigma^2 I` only | any positive-definite `V` |
| Trial design vocabulary | none | phases, accrual, dropout, interim |
| Fitting | comprehensive | one reference fitter; `lme4` plugs in |

Use `lme4` to fit. Use `zzrctsim` to generate, drive the replicates,
and measure. The overlap in generation is real but narrow, and where
the covariance structure is the object of study rather than a
nuisance, the constraint above decides it.
