---
title: "Two ways to declare a data-generating mechanism"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two ways to declare a data-generating mechanism}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzrctsim)
```

## The choice

A longitudinal data-generating mechanism can be written two ways.

**Conditionally**, through random effects: draw `b_i ~ N(0, G)`, then
`y_i = X_i beta + Z_i b_i + e_i`. This is what `lme4`, `simr` and
`glmmTMB` take.

**Marginally**, through the covariance of the response vector `V`
directly. This is what `mmrm` and `nlme::gls` take.

Most packages accept one or the other. `zzrctsim` accepts either, and
converts between them where a conversion exists.

## Conditional to marginal is always exact

```{r cond}
tt <- 1:6
g <- dgm_conditional(G = diag(c(9, 0.04)), sigma2 = 4)
round(cov_at(g, tt), 2)
```

The induced covariance is `Z G Z' + sigma^2 I`, computed exactly.
`as_marginal()` freezes it into a marginal DGM on a stated grid.

## Marginal to conditional is not always possible

Declare a covariance directly. `cov_ar1()` and `cov_cs()` return
covariance *functions*, so they extend to any grid.

```{r marg}
ar1 <- cov_ar1(sd = 1, rho = 0.6)
round(ar1(tt), 3)
```

Now ask whether a random-effects model could have produced this.

```{r certify}
certify(ar1(tt))
```

`q_min = 5` means AR(1) at six visits needs **five** random effects.
Compound symmetry, by contrast, needs one:

```{r certify-cs}
certify(cov_cs(sd = 1, rho = 0.5)(tt))$q_min
```

That single number explains a fact practitioners meet constantly: a
random intercept model produces compound symmetry and nothing else.

## Why

`V = Z G Z' + sigma^2 I` requires `Z G Z'` to be positive
semi-definite with rank at most `q`. Subtracting `sigma^2 I` shifts
every eigenvalue equally, so the question is settled by the spectrum.

```{r spectrum}
round(eigen(ar1(tt))$values, 4)
```

The eigenvalues are distinct, so one common downward shift can zero at
most one of them. Take `sigma^2` to be the smallest and five positive
eigenvalues remain; take it any larger and the matrix stops being
positive semi-definite. Hence `q >= 5`.

For compound symmetry the smallest eigenvalue has multiplicity `p - 1`,
so a single shift zeroes all of them at once:

```{r spectrum-cs}
round(eigen(cov_cs(sd = 1, rho = 0.5)(tt))$values, 4)
```

## The certificate is constructive

It does not merely answer yes or no. It hands back the construction.

```{r reconstruct}
ct <- certify(ar1(tt))
max(abs(reconstruct(ct) - ar1(tt)))
```

Exact to machine precision.

## Refusing rather than approximating

`as_conditional()` will not quietly return an approximation when the
budget is too small.

```{r refuse, error = TRUE}
as_conditional(dgm_marginal(ar1), tt, q = 2)
```

If you want to know how wrong a restricted budget would be,
`reach_error()` says so:

```{r reach-error}
reach_error(ar1(tt))
```

A single random effect is wrong by 28% in relative Frobenius norm.
That is not an approximation to AR(1); it is a different covariance.

## Two qualifications

**`q_min` is necessary, not sufficient for a particular model.** The
certificate allows `Z` to be the eigenvectors of `V`. A real
random-effects specification fixes `Z` to design columns — an
intercept, a slope in time. Compare the leading eigenvectors:

```{r eigenvectors}
round(certify(cov_cs(1, 0.5)(tt))$Z[, 1], 3)
round(certify(ar1(tt))$Z[, 1], 3)
```

Compound symmetry's is flat, which is exactly the all-ones direction a
random intercept is confined to — hence the perfect match. AR(1)'s is
bowed, so even the best one-dimensional direction is unavailable to a
random intercept. The certificate proves impossibility cleanly; it
proves possibility only for an unconstrained `Z`.

**The mathematics is classical.** Writing `V = Lambda Lambda' +
sigma^2 I` and solving by eigendecomposition is probabilistic PCA,
equivalently factor analysis with isotropic uniquenesses. What is
useful here is not the algebra but having it run automatically, so
that a user who supplies an empirically estimated covariance is told
at once whether the random-effects analysis they intend can represent
it at all.

## Why this matters in practice

Suppose you estimated this covariance from a completed trial —
heterogeneous variances rising over follow-up, correlations decaying
with lag:

```{r empirical}
sds <- c(4.5, 4.5, 5.4, 6.3, 7.2, 8.2)
V_emp <- cov_ar1(sd = sds, rho = 0.6)(tt)
certify(V_emp)$q_min
```

Five random effects for six visits. A study that plans to generate
from this covariance and analyse with a random intercept-and-slope
model is generating something its analysis model cannot represent.
That may be exactly the misspecification you want to study — but it
should be a choice, not a surprise.

Generating from it is unaffected: the marginal route needs no
random-effects representation.

```{r generate-marginal}
set.seed(2)
s <- trial_schedule(treatment = 5, interval = 3)
arm <- factor(rep(c("placebo", "active"), each = 50),
              levels = c("placebo", "active"))
d <- runin_design(s, arm, reference = "placebo")
gm <- dgm_marginal(cov_ar1(sd = sds[1:6], rho = 0.6))
dat <- generate_outcomes(d, gm,
                         beta = c(x_slope = 0.5, x_trt_active = -0.25),
                         intercept = 20)
round(cor(matrix(dat$y, ncol = nrow(s), byrow = TRUE))[1, ], 3)
```
