---
title: "Getting started with zzrctsim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with zzrctsim}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzrctsim)
```

## The five steps

A simulation study in `zzrctsim` is always the same five steps, and
they are separate on purpose: each corresponds to one element of
ADEMP, so a manuscript's Methods subsections map onto code rather than
having to be reconstructed from a script.

1. **Design** — when are subjects seen?
2. **Data-generating mechanism** — what produces the response?
3. **Generation** — draw one trial.
4. **Analysis** — fit something, returning a standard shape.
5. **Performance** — summarize across replicates, with Monte Carlo
   standard errors.

## 1. Design

`trial_schedule()` describes the visit times. The simplest case is a
fixed number of post-randomization visits at a regular interval.

```{r schedule}
s <- trial_schedule(treatment = 4, interval = 3)
s
```

The randomization visit is at `time = 0`. `runin_design()` expands the
schedule over subjects and builds the model columns.

```{r design}
arm <- factor(rep(c("placebo", "active"), each = 100),
              levels = c("placebo", "active"))
d <- runin_design(s, arm, reference = "placebo")
head(d, 3)
```

Three columns are produced: `x_slope` is the common slope on every
visit, `x_hinge` is the reference arm's change of slope at
randomization, and `x_trt_active` is the treatment effect on the rate
of change. Their coefficients are what you supply next.

## 2. Data-generating mechanism

A DGM can be declared through random effects or through a covariance
matrix. Here is the random-effects form: a random intercept with
variance 9, a random slope with variance 0.04, and residual variance
4.

```{r dgm}
g <- dgm_conditional(G = diag(c(9, 0.04)), sigma2 = 4)
round(cov_at(g, s$time), 2)
```

`cov_at()` shows the covariance this implies for the response. Note it
is a function of the visit times, not a fixed matrix — which is what
lets subjects sit on different grids when accrual is staggered.

## 3. Generation

```{r generate}
set.seed(1)
b <- c(x_slope = 0.5, x_trt_active = -0.25)
dat <- generate_outcomes(d, g, beta = b, intercept = 20)
head(dat[, c("id", "arm", "time", "mu", "y")], 4)
```

`mu` is the fixed-effect mean and `y` the realized response. The data
are complete at this point; missingness is a separate step, which is
what makes it possible to measure how much dropout costs.

## 4. Analysis

Any analysis method may be used, provided it returns a `fit_result()`.
`fit_ancova()` is supplied as a reference: change from baseline at the
final visit, adjusted for the baseline value.

```{r analyze}
fit_ancova(dat, reference = "placebo")
```

Writing your own fitter is a matter of extracting three numbers:

```{r custom-fitter}
fit_lm_final <- function(z) {
  zz <- z[z$time == max(z$time), ]
  m <- stats::lm(y ~ arm, data = zz)
  cf <- summary(m)$coefficients
  fit_result(estimate = cf[2, "Estimate"],
             se = cf[2, "Std. Error"],
             p_value = cf[2, "Pr(>|t|)"],
             df = m$df.residual)
}
fit_lm_final(dat)$estimate
```

## 5. Replicates and performance

`run_simulation()` drives the loop. The estimand is declared
explicitly so that bias and coverage have a referent.

```{r simulate}
theta <- b[["x_trt_active"]] * max(s$time)
res <- run_simulation(
  B = 200,
  generate = function() generate_outcomes(d, g, beta = b,
                                          intercept = 20),
  analyze = list(ancova = function(z)
    fit_ancova(z, reference = "placebo")),
  estimand = estimand("final-visit contrast", theta),
  seed = 42L
)
res
```

```{r performance}
compute_performance(res)
```

Every measure carries a Monte Carlo standard error. That is what tells
you whether an apparent bias is real or is noise from too few
replicates. `nsim_for_mcse()` inverts the relationship if you would
rather choose `B` from a precision requirement:

```{r nsim}
nsim_for_mcse("proportion", mcse = 0.01, p = 0.8)
```

## Comparing methods on the same data

Passing several fitters analyzes the *identical* data set with each,
which is the paired comparison Morris et al. recommend. The estimates
are then correlated, and the comparison is not contaminated by
between-dataset variation.

```{r crn}
res2 <- run_simulation(
  B = 100,
  generate = function() generate_outcomes(d, g, beta = b,
                                          intercept = 20),
  analyze = list(
    ancova = function(z) fit_ancova(z, reference = "placebo"),
    final_only = fit_lm_final),
  estimand = estimand("final-visit contrast", theta),
  seed = 7L
)
perf <- compute_performance(res2)
perf[perf$measure %in% c("bias", "emp_se", "coverage"), ]
```

ANCOVA has the smaller empirical standard error, because adjusting for
baseline removes variation the unadjusted comparison leaves in.

## Where to go next

- `vignette("dual-parameterization")` — declaring a DGM through a
  covariance instead of random effects, and when the two are
  interchangeable.
- `vignette("missing-data")` — dropout mechanisms, reusable masks, and
  reference-based trajectories.
- `vignette("time-to-event")` — single and recurrent events.
- `vignette("comparison-lme4")` and `vignette("comparison-simr")` —
  how this relates to tools you may already use.
