---
title: "Missing data: mechanisms, masks, and reference-based trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing data: mechanisms, masks, and reference-based trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzrctsim)
set.seed(11)
```

## Complete data first

Missingness is always applied *after* complete data are generated. That
ordering is not incidental. It is what lets you measure how much
dropout costs, by comparing against the complete-data run, and it is
the only way a mechanism that depends on the response can be expressed
at all — a hazard that reads `y` needs `y` to exist.

```{r data}
s <- trial_schedule(treatment = 5, interval = 3)
arm <- factor(rep(c("placebo", "active"), each = 300),
              levels = c("placebo", "active"))
d <- runin_design(s, arm, reference = "placebo")
g <- dgm_conditional(G = diag(c(9, 0.04)), sigma2 = 4)
b <- c(x_slope = 0.5, x_trt_active = -0.25)
dat <- generate_outcomes(d, g, beta = b, intercept = 20)
mean(is.na(dat$y))
```

## One model, three mechanisms

The dropout hazard follows Diggle and Kenward (1994). For a subject
still under observation at visit *j*:

```
logit P(drop at j) = psi0 + psi1 (y_{j-1} - c) + psi2 (y_j - c)
```

The three mechanisms are **nested restrictions of one model**, not
three separate models:

```{r mechanisms}
c(dropout_mechanism(0, 0),
  dropout_mechanism(0.2, 0),
  dropout_mechanism(0.2, 0.1))
```

MNAR therefore *extends* MAR — setting `psi2` does not remove the
dependence on history. That matters for sensitivity analysis: a
contrast between MAR and MNAR should vary one thing, not two.

## The target is the realized dropout

`psi0` is calibrated so that the requested proportion is the one you
get, whatever `psi1` and `psi2` are. Without this, a hazard specified
on an untransformed scale saturates:

```{r saturation}
# An uncentred hazard at a realistic ADAS-Cog level:
plogis(qlogis(0.12) + 0.25 * 20)
```

Every subject would drop out at the first eligible visit while the
nominal rate still read 0.12. `zzrctsim` centers the response and
solves for `psi0`:

```{r calibration}
spec <- function(target, psi1 = 0, psi2 = 0) {
  m <- dropout_mask(dat, target = target, psi1 = psi1, psi2 = psi2)
  sp <- attr(m, "spec")
  data.frame(mechanism = sp$mechanism, target = target,
             psi0 = round(as.numeric(sp$psi0), 3),
             expected = round(as.numeric(sp$expected), 4),
             realized = round(mean(tapply(m$missing, m$id, any)), 4))
}
rbind(spec(0.15), spec(0.30),
      spec(0.30, psi1 = 0.15),
      spec(0.30, psi1 = 0.15, psi2 = 0.10))
```

`psi0` moves to absorb the change; the target does not.

Because the hazards depend only on complete-data responses, the
expected proportion has a closed form, `1 - prod(1 - p_j)` averaged
over subjects. Calibration is therefore a one-dimensional root-find
with no inner Monte Carlo.

## Differential dropout by arm

Real trials lose more subjects from one arm than the other, and for
compendium 14 this is the whole subject. Supply a named target and a
grouping column:

```{r by-arm}
dat$armf <- arm[dat$id]
m_by <- dropout_mask(dat, target = c(placebo = 0.15, active = 0.35),
                     by = "armf")
m_by
ever <- tapply(m_by$missing, m_by$id, any)
round(tapply(ever, arm[as.integer(names(ever))], mean), 3)
```

`psi_cov` adds covariate-dependent terms in the same way.

## Monotone and intermittent

Dropout is absorbing; intermittent missingness is not. Both calibrate
to the same target, because the target is "any missing observation".

```{r intermittent}
mono <- dropout_mask(dat, target = 0.30, monotone = TRUE)
intr <- dropout_mask(dat, target = 0.30, monotone = FALSE)
data.frame(
  kind = c("monotone", "intermittent"),
  observations_lost = c(sum(mono$missing), sum(intr$missing)),
  subjects_affected = c(length(unique(mono$id[mono$missing])),
                        length(unique(intr$id[intr$missing]))))
```

The same subjects are affected — that is what calibration fixes — but
absorbing dropout costs far more data.

## Masks are objects

Generation and application are separate layers. A mask can be saved,
inspected, imported from a completed trial, and **reused across
replicates**, which is what holds the missingness pattern fixed under
common random numbers.

```{r masks}
masked <- apply_mask(dat, mono)
sum(is.na(masked$y))

# The same pattern applied to an independent replicate:
dat2 <- generate_outcomes(d, g, beta = b, intercept = 20)
masked2 <- apply_mask(dat2, mono)
identical(is.na(masked$y), is.na(masked2$y))
```

An empirical pattern from a real trial is imported with
`mask_from_data()`, which also detects whether it is monotone:

```{r import}
back <- mask_from_data(masked)
c(identical(back$missing, mono$missing),
  attr(back, "spec")$monotone)
```

## Reference-based trajectories

`rbmi` implements reference-based assumptions for *analysis*. Here they
are **generators**: they produce data whose true post-discontinuation
behavior follows a named assumption, so that an analysis assuming
something else can be measured for bias.

```{r refbased}
mk <- dropout_mask(dat, target = 0.40)
shift <- vapply(c("j2r", "cir", "lmcf", "cr"), function(method) {
  z <- reference_based(dat, mk, method = method)
  post <- z$discontinued & dat$armf == "active"
  mean(z$mu[post] - dat$mu[post])
}, numeric(1))
round(shift, 3)
```

The ordering `0 < CIR < J2R` is the expected one: copying increments
in the reference retains the accumulated benefit that jumping to the
reference discards. Each subject's realized residual is preserved, so
within-subject correlation survives the shift.

**Discontinuation is not missingness.** `reference_based()` changes
the trajectory; `apply_mask()` hides observations. They are separate
calls because under a treatment-policy estimand post-discontinuation
values are observed, and under a hypothetical estimand they are not.
Fusing them would make the treatment-policy case — the one ICH E9(R1)
puts first — inexpressible.

## What dropout costs

The reason for keeping the layers separate: you can price the
missingness directly.

```{r cost}
run <- function(drop) {
  gen <- function() {
    z <- generate_outcomes(d, g, beta = b, intercept = 20)
    if (drop) z <- apply_mask(z, dropout_mask(z, target = 0.30,
                                              psi1 = 0.10))
    z
  }
  res <- run_simulation(
    B = 150, generate = gen,
    analyze = list(a = function(z) fit_ancova(z, reference = "placebo")),
    estimand = estimand("final contrast",
                        b[["x_trt_active"]] * max(s$time)),
    seed = 4L)
  p <- compute_performance(res)
  p$estimate[p$measure %in% c("emp_se", "rejection")]
}
data.frame(scenario = c("complete", "30% MAR dropout"),
           rbind(run(FALSE), run(TRUE)))
```

The columns are empirical standard error and power. Dropout inflates
the first and reduces the second; `sample_size()` prices the same
thing in subjects.
