---
title: "Time-to-event endpoints: single, recurrent, and derived"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-to-event endpoints: single, recurrent, and derived}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzrctsim)
has_surv <- requireNamespace("survival", quietly = TRUE)
knitr::opts_chunk$set(eval = TRUE)
```

## Why this is a separate generator

Time-to-event does not fit the covariance-based parameterization used
for continuous outcomes, and that is structural rather than an
oversight. Censoring, hazards and risk sets are a different kind of
object from a marginal law, so `dgm_tte()` sits beside
`dgm_conditional()` rather than being another `family` argument.

Generation is by **inversion of the cumulative hazard**, which is
exact. Under proportional hazards the individual cumulative hazard is
`H_i(t) = H_0(t) exp(lp_i)`, so if `E ~ Exp(1)` then
`T = H_0^{-1}(E / exp(lp_i))` has precisely the intended
distribution — no grid, no discretization.

## Single event

```{r single}
set.seed(4001)
subj <- data.frame(id = 1:4000, trt = rep(0:1, each = 2000))
g <- dgm_tte(shape = 1.6, scale = 15)
g
ev <- generate_tte(subj, g, beta = c(trt = log(0.65)), followup = 30)
head(ev, 3)
```

`status` is 1 for an event and 0 for administrative censoring. Checking
that the mechanism is what it claims:

```{r single-check, eval = has_surv}
fit <- survival::coxph(survival::Surv(time, status) ~ trt, data = ev)
c(true_log_HR = log(0.65),
  estimated = unname(coef(fit)),
  se = sqrt(unname(vcov(fit)[1, 1])),
  censored = mean(ev$status == 0))
```

## Follow-up from staggered accrual

`followup` may be a vector, which is what makes this compose with a
common close-out date: subjects enrolled early are followed longer.

```{r accrual}
set.seed(12)
enrol <- accrue(nrow(subj), period = 12, pattern = "uniform")
close <- max(enrol) + 24
ev2 <- generate_tte(subj, g, beta = c(trt = log(0.65)),
                    followup = close - enrol)
round(quantile(close - enrol, c(0, 0.5, 1)), 2)
mean(ev2$status)
```

Follow-up now ranges from 24 months for the last enrollee up to 36 for
the first, and the event rate reflects that.

## Baseline hazards

The Weibull covers monotone hazards and reduces to the exponential at
`shape = 1`. The **piecewise exponential** expresses an arbitrary
hazard shape without committing to a parametric family, which is why
it is the usual choice in trial simulation.

```{r piecewise}
set.seed(4101)
g_pw <- dgm_tte(breaks = c(6, 12), rates = c(0.02, 0.06, 0.15))
g_pw
sp <- data.frame(id = 1:40000, trt = 0)
e_pw <- generate_tte(sp, g_pw, beta = c(trt = 0), followup = 20)
emp <- function(lo, hi) {
  sum(e_pw$status == 1L & e_pw$time > lo & e_pw$time <= hi) /
    sum(pmax(0, pmin(e_pw$time, hi) - lo))
}
data.frame(interval = c("[0,6)", "[6,12)", "[12,20]"),
           specified = c(0.02, 0.06, 0.15),
           realized = round(c(emp(0, 6), emp(6, 12), emp(12, 20)), 4))
```

## A delayed treatment effect

`effect_lag` breaks proportional hazards deliberately: the arms are
indistinguishable until the lag and diverge after it. This is the
shape immuno-oncology trials show, and the one anti-amyloid trials in
Alzheimer's disease are generally argued to show. A constant hazard
ratio cannot represent it.

```{r lag, eval = has_surv}
set.seed(4102)
s4 <- data.frame(id = 1:12000, trt = rep(0:1, each = 6000))
e_lag <- generate_tte(s4, dgm_tte(shape = 1, scale = 20,
                                  effect_lag = 8),
                      beta = c(trt = log(0.4)), followup = 40)
fz <- survival::coxph(survival::Surv(time, status) ~ trt, data = e_lag)
data.frame(true_post_lag_HR = 0.4,
           fitted_constant_HR = round(exp(unname(coef(fz))), 3),
           ph_test_p = signif(survival::cox.zph(fz)$table[1, "p"], 3))
```

The fitted constant hazard ratio is attenuated toward one, and
`cox.zph()` rejects proportionality. Both are intended: the generator
produces data a proportional-hazards summary misrepresents, which is
what you need in order to quantify how badly.

## Recurrent events

Setting `recurrent = TRUE` generates all events in the follow-up
window, returned in counting-process format — one row per interval,
which is what `coxph()` expects for an Andersen-Gill model.

```{r recurrent}
set.seed(4002)
subj2 <- data.frame(id = 1:1500, trt = rep(0:1, each = 750))
g_rec <- dgm_tte(shape = 1, scale = 10, recurrent = TRUE)
rec <- generate_tte(subj2, g_rec, beta = c(trt = 0), followup = 20)
head(rec, 4)
```

The intervals for a subject tile the window contiguously and end with
a censored row.

A Gaussian **frailty** on the log-hazard scale induces within-subject
dependence between gap times, which is what distinguishes a recurrent
process from independent replications. Its signature is
overdispersion:

```{r frailty}
set.seed(4003)
counts <- function(sd) {
  gg <- dgm_tte(shape = 1, scale = 10, recurrent = TRUE,
                frailty_sd = sd)
  r <- generate_tte(subj2, gg, beta = c(trt = 0), followup = 20)
  unlist(tapply(r$status, r$id, sum))
}
data.frame(frailty_sd = c(0, 0.8),
           mean = round(c(mean(counts(0)), mean(counts(0.8))), 3),
           variance = round(c(var(counts(0)), var(counts(0.8))), 3))
```

Without frailty the counts are Poisson, variance equal to mean. With
frailty they are overdispersed.

A frailty also makes the *marginal* distribution non-proportional even
though the conditional model is proportional hazards. That is the
usual reason a fitted hazard ratio attenuates toward one, and it means
an `estimand()` must say whether it names the conditional or the
marginal hazard ratio.

## An endpoint derived from a longitudinal trajectory

Comparing MMRM against survival analysis is only meaningful if both
endpoints describe the *same* underlying process rather than being
drawn independently. `tte_from_trajectory()` defines the event as the
first visit at which a generated trajectory crosses a threshold.

```{r derived}
set.seed(4005)
s6 <- trial_schedule(treatment = 6, interval = 3)
arm6 <- factor(rep(c("placebo", "active"), each = 300),
               levels = c("placebo", "active"))
d6 <- runin_design(s6, arm6, reference = "placebo")
lo <- generate_outcomes(d6, dgm_conditional(G = diag(c(4, 0.02)),
                                            sigma2 = 2),
                        beta = c(x_slope = 0.35, x_trt_active = -0.20),
                        intercept = 10)
evd <- tte_from_trajectory(lo, threshold = 16, direction = "above")
a6 <- arm6[evd$id]
data.frame(arm = c("placebo", "active"),
           crossed = round(c(mean(evd$status[a6 == "placebo"]),
                             mean(evd$status[a6 == "active"])), 3),
           mean_time_given_crossed = round(c(
             mean(evd$time[evd$status == 1 & a6 == "placebo"]),
             mean(evd$time[evd$status == 1 & a6 == "active"])), 2))
```

This table repays a second look. The treatment effect is large — 64%
of placebo cross against 25% of active — yet the mean crossing time
*among those who crossed* is nearly identical between arms.

Conditioning on having crossed selects the fastest progressors within
each arm, and selects more severely in the arm where crossing is
rarer. The correct comparison retains the censored observations:

```{r derived-cox, eval = has_surv}
fd <- survival::coxph(survival::Surv(time, status) ~ a6, data = evd)
c(hazard_ratio = round(exp(unname(coef(fd))), 3),
  p = signif(summary(fd)$coefficients[1, "Pr(>|z|)"], 3))
```

It is a compact demonstration of why a survival endpoint is analyzed
through the hazard rather than through the mean event time of the
subset that had events.

Subjects who never cross are censored at their last observed visit, so
dropout and administrative censoring both propagate into the survival
endpoint correctly.

## What is not implemented

Accelerated failure time models, competing risks, cure models, and
baseline hazards other than Weibull or piecewise exponential. For a
trial where death competes with progression — common in Alzheimer's
disease — treating death as censoring is the classic error, and this
package currently gives you no way to avoid it.
